---
title: "MPRA: promoter activity"
subtitle: "Barcode-level RNA/DNA ratios summarised to promoters"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4.5, dpi = 150)
```

Activity is computed per barcode as

$$\text{activity} = \log_2\!\left(\frac{\text{RNA}_{\text{CPM}}}{\text{pDNA}_{\text{CPM}}} + \varepsilon\right)$$

and averaged over the barcodes of each promoter. Where two conditions are
present, `Delta = condB - condA` is the difference in promoter activity between
them.

# Setup

```{r libs}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(tibble)
  library(ggplot2); library(patchwork); library(ggrepel)
  library(DESeq2)
})
set.seed(42)

# Bioconductor packages mask dplyr verbs; make dplyr win for this session.
filter <- dplyr::filter; select <- dplyr::select; count <- dplyr::count
rename <- dplyr::rename; slice  <- dplyr::slice; n_distinct <- dplyr::n_distinct
```

## Parameters — edit these

```{r params}
results_dir <- "../../MPRA/results"      # output of workflow/04_build_matrix
pba_path    <- "../../PBA/results/PBA-lib-rep1/pba/PBA-lib-rep1.pba_clean.tsv"
out_dir     <- "results"

use_starcode  <- TRUE
starcode_dist <- 2

pdna_min        <- 8    # min reads per barcode in EACH pDNA input
min_bc_per_prom <- 5    # min barcodes for a promoter to be quantified
delta_cut       <- 1    # |Delta| threshold for calling a difference

samples <- read_tsv("../config/samples_mpra.tsv", show_col_types = FALSE) %>%
  mutate(key = case_when(type == "pDNA" ~ paste0("pDNA_", condition),
                         TRUE ~ paste(condition, replicate, sep = "_")))
rna_keys  <- samples$key[samples$type == "RNA"]
pdna_keys <- samples$key[samples$type == "pDNA"]

# Condition levels: the first is the reference in every comparison.
cond_levels <- c("condA", "condB")
samples
```

```{r theme}
fig_dir <- file.path(out_dir, "figures"); table_dir <- file.path(out_dir, "tables")
for (d in c(fig_dir, table_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

theme_pub <- function(base_size = 11) {
  theme_classic(base_size = base_size) +
    theme(panel.grid.major = element_line(color = "grey92", linewidth = 0.3),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold"),
          plot.title = element_text(face = "bold", size = base_size + 1))
}
cond_cols   <- setNames(c("#4C72B0", "#C44E52"), cond_levels)
effect_cols <- c(Lower = "#08519C", Unchanged = "grey70", Higher = "#A50F15")

revcomp <- function(x) {
  x <- chartr("ACGTNacgtn", "TGCANtgcan", x)
  vapply(strsplit(x, ""), function(z) paste(rev(z), collapse = ""), character(1))
}
cpm <- function(x) x / sum(x) * 1e6
```

# Count matrix

```{r load-matrix}
mat_file <- if (use_starcode)
  sprintf("library_count_matrix.starcode_d%d.tsv", starcode_dist) else
  "library_count_matrix.tsv"
mat_path <- file.path(results_dir, "aggregated", mat_file)
stopifnot(file.exists(mat_path))

cnt_raw <- read_tsv(mat_path, show_col_types = FALSE)
present <- intersect(samples$sample_id, names(cnt_raw))
if (length(setdiff(samples$sample_id, present)))
  warning("Missing from matrix: ",
          paste(setdiff(samples$sample_id, present), collapse = ", "))

cnt_mat <- as.matrix(cnt_raw[, present, drop = FALSE])
rownames(cnt_mat) <- cnt_raw$barcode
colnames(cnt_mat) <- samples$key[match(present, samples$sample_id)]
storage.mode(cnt_mat) <- "integer"
cat(nrow(cnt_mat), "barcodes x", ncol(cnt_mat), "samples\n")
```

## Read composition

```{r qc-composition, fig.width=8, fig.height=4}
qc <- read_tsv(file.path(results_dir, "aggregated", "qc_summary_all.tsv"),
               show_col_types = FALSE) %>%
  rename(sample_id = sample) %>%
  left_join(samples, by = "sample_id") %>%
  mutate(assigned = library_reads,
         unassigned = total_reads - library_reads) %>%
  select(key, type, total_reads, assigned, unassigned)
write_tsv(qc, file.path(table_dir, "read_composition.tsv"))

p_qc <- qc %>%
  pivot_longer(c(assigned, unassigned),
               names_to = "category", values_to = "reads") %>%
  ggplot(aes(key, reads, fill = category)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c(assigned = "#08306B", unassigned = "grey80")) +
  labs(x = NULL, y = "Fraction of reads", title = "Read composition") +
  theme_pub() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_qc
ggsave(file.path(fig_dir, "01_read_composition.pdf"), p_qc, width = 8, height = 4)
```

# Promoter–barcode assignment

```{r load-pba}
pba <- read_tsv(pba_path, show_col_types = FALSE) %>%
  select(promoter, barcode) %>% distinct()

# Orientation check against the count matrix.
if (length(intersect(rownames(cnt_mat), revcomp(pba$barcode))) >
    length(intersect(rownames(cnt_mat), pba$barcode))) {
  pba$barcode <- revcomp(pba$barcode)
  message("Using reverse-complemented PBA barcodes.")
}

pba <- pba %>%
  group_by(barcode) %>% filter(n_distinct(promoter) == 1) %>% ungroup() %>%
  mutate(kind = case_when(grepl("^R[0-9]", promoter) ~ "scrambled_control",
                          grepl("^P[0-9]", promoter) ~ "test_promoter",
                          TRUE ~ "other_control"))
cat(n_distinct(pba$barcode), "barcodes,", n_distinct(pba$promoter), "promoters\n")
count(distinct(pba, promoter, kind), kind)
```

# Filtering

Keep PBA barcodes seen at least `r pdna_min` times in **each** pDNA input.

```{r filter}
bc_anno <- as_tibble(cnt_mat, rownames = "barcode") %>%
  inner_join(pba, by = "barcode") %>%
  filter(if_all(all_of(pdna_keys), ~ .x >= pdna_min))

cnt_flt <- as.matrix(bc_anno[, samples$key]); rownames(cnt_flt) <- bc_anno$barcode
cat(nrow(bc_anno), "barcodes,", n_distinct(bc_anno$promoter), "promoters retained\n")
```

# Normalization and barcode activity

Each sample is converted to counts per million over the retained barcodes, so
sequencing depth cancels. The pDNA aliquots come from the same plasmid pool and
are averaged into one input reference.

```{r activity}
rna_cpm  <- apply(cnt_flt[, rna_keys,  drop = FALSE], 2, cpm)
pdna_cpm <- rowMeans(apply(cnt_flt[, pdna_keys, drop = FALSE], 2, cpm))

ratio <- sweep(rna_cpm, 1, pdna_cpm, "/")

# One shared pseudocount (10th percentile of non-zero ratios) so the floor is
# identical across samples and cancels when conditions are subtracted.
eps <- quantile(ratio[ratio > 0], 0.1, names = FALSE)
cat("Pseudocount:", signif(eps, 3), "\n")

bc_act <- log2(ratio + eps) %>%
  as.data.frame() %>% rownames_to_column("barcode") %>%
  pivot_longer(-barcode, names_to = "sample", values_to = "activity") %>%
  separate(sample, c("condition", "replicate"), sep = "_") %>%
  left_join(select(bc_anno, barcode, promoter, kind), by = "barcode")
summary(bc_act$activity)
```

# Promoter-level activity

```{r promoter-level}
valid_prom <- bc_anno %>% count(promoter) %>%
  filter(n >= min_bc_per_prom) %>% pull(promoter)

prom_wide <- bc_act %>%
  filter(promoter %in% valid_prom) %>%
  group_by(promoter, kind, condition, replicate) %>%
  summarise(activity = mean(activity, na.rm = TRUE), .groups = "drop") %>%
  unite("sample", condition, replicate) %>%
  pivot_wider(names_from = sample, values_from = activity)
cat(nrow(prom_wide), "promoters with >=", min_bc_per_prom, "barcodes\n")
```

```{r replicate-correlation, fig.width=8, fig.height=4}
pair_plot <- function(d, x, y) {
  d <- drop_na(select(d, all_of(c(x, y))))
  ggplot(d, aes(.data[[x]], .data[[y]])) +
    geom_point(alpha = 0.4, size = 0.8, color = "#08306B") +
    geom_abline(linetype = 2, color = "grey50") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3, size = 3.2,
             label = sprintf("r=%.3f", cor(d[[x]], d[[y]]))) +
    labs(x = x, y = y) + theme_pub(10)
}
p_rep <- pair_plot(prom_wide, paste0(cond_levels[1], "_rep1"),
                              paste0(cond_levels[1], "_rep2")) |
         pair_plot(prom_wide, paste0(cond_levels[2], "_rep1"),
                              paste0(cond_levels[2], "_rep2"))
p_rep
ggsave(file.path(fig_dir, "02_replicate_correlation.pdf"), p_rep, width = 8, height = 4)
```

## Activity per promoter

```{r summarise}
prom_activity <- prom_wide %>%
  mutate(A = rowMeans(across(starts_with(paste0(cond_levels[1], "_")))),
         B = rowMeans(across(starts_with(paste0(cond_levels[2], "_")))),
         delta = B - A,
         effect = factor(case_when(delta <= -delta_cut ~ "Lower",
                                   delta >=  delta_cut ~ "Higher",
                                   TRUE ~ "Unchanged"),
                         levels = names(effect_cols))) %>%
  rename(!!cond_levels[1] := A, !!cond_levels[2] := B)

pa <- filter(prom_activity, kind == "test_promoter")
count(pa, effect) %>% mutate(pct = round(100 * n / sum(n), 1))
write_tsv(prom_activity, file.path(table_dir, "promoter_activity.tsv"))
```

```{r figs-activity, fig.width=10, fig.height=4.5}
p_rank <- prom_activity %>%
  arrange(.data[[cond_levels[1]]]) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(rank, .data[[cond_levels[1]]], color = kind)) +
  geom_point(size = 1, alpha = 0.7) +
  labs(x = "Promoter (ranked)", y = "Activity", title = "Promoter activity") +
  theme_pub()

p_sc <- ggplot(pa, aes(.data[[cond_levels[1]]], .data[[cond_levels[2]]],
                       color = effect)) +
  geom_abline(linetype = 2, color = "grey50") +
  geom_abline(slope = 1, intercept = c(-delta_cut, delta_cut),
              linetype = 3, color = "grey70") +
  geom_point(size = 2, alpha = 0.6) +
  geom_text_repel(data = slice_max(pa, abs(delta), n = 10),
                  aes(label = promoter), size = 2.5, color = "black",
                  max.overlaps = 15) +
  scale_color_manual(values = effect_cols) +
  coord_equal() + theme_pub()

p_rank | p_sc
ggsave(file.path(fig_dir, "03_promoter_activity.pdf"), p_rank | p_sc,
       width = 10, height = 4.5)
```

# Differential activity between conditions

DESeq2 on per-promoter summed RNA counts, with default median-of-ratios size
factors. Drop this section if you only need the activity table above.

```{r deseq2}
prom_counts <- bc_anno %>%
  filter(promoter %in% valid_prom) %>%
  group_by(promoter) %>%
  summarise(across(all_of(samples$key), sum), .groups = "drop")
rna_cnt <- as.matrix(prom_counts[, rna_keys]); rownames(rna_cnt) <- prom_counts$promoter

coldata <- samples %>% filter(type == "RNA") %>%
  arrange(match(key, colnames(rna_cnt))) %>%
  mutate(condition = factor(condition, levels = cond_levels),
         replicate = factor(replicate)) %>%
  as.data.frame()
rownames(coldata) <- coldata$key

dds <- DESeqDataSetFromMatrix(rna_cnt, coldata, design = ~ replicate + condition)
dds <- DESeq(dds)
res <- results(dds, name = sprintf("condition_%s_vs_%s",
                                   cond_levels[2], cond_levels[1]), alpha = 0.05)

res_df <- as.data.frame(res) %>% rownames_to_column("promoter") %>% arrange(padj)
write_tsv(res_df, file.path(table_dir, "differential_activity.tsv"))
summary(res)
```

```{r volcano, fig.width=6.5, fig.height=5}
p_vol <- res_df %>%
  mutate(sig = case_when(is.na(padj) | padj >= 0.05 ~ "NS",
                         log2FoldChange <= -delta_cut ~ "Lower",
                         log2FoldChange >=  delta_cut ~ "Higher",
                         TRUE ~ "Sig, small effect")) %>%
  ggplot(aes(log2FoldChange, -log10(padj), color = sig)) +
  geom_vline(xintercept = c(-delta_cut, delta_cut), linetype = 2, color = "grey70") +
  geom_hline(yintercept = -log10(0.05), linetype = 2, color = "grey70") +
  geom_point(alpha = 0.7, size = 1.2) +
  scale_color_manual(values = c(Lower = "#08519C", Higher = "#A50F15",
                                `Sig, small effect` = "#FDB863", NS = "grey80")) +
  labs(x = expression(log[2]~fold~change), y = expression(-log[10]~adj.~p)) +
  theme_pub()
p_vol
ggsave(file.path(fig_dir, "04_volcano.pdf"), p_vol, width = 6.5, height = 5)
```

Barcodes are shared between conditions, so a paired test over barcodes is an
independent check on DESeq2.

```{r paired-test}
paired <- bc_act %>%
  group_by(promoter, barcode, condition) %>%
  summarise(activity = mean(activity), .groups = "drop") %>%
  pivot_wider(names_from = condition, values_from = activity) %>%
  drop_na(all_of(cond_levels)) %>%
  group_by(promoter) %>% filter(n() >= min_bc_per_prom) %>%
  summarise(n_bc = n(),
            median_delta = median(.data[[cond_levels[2]]] - .data[[cond_levels[1]]]),
            p = tryCatch(wilcox.test(.data[[cond_levels[2]]],
                                     .data[[cond_levels[1]]], paired = TRUE)$p.value,
                         error = function(e) NA_real_),
            .groups = "drop") %>%
  mutate(padj = p.adjust(p, "BH"))
write_tsv(paired, file.path(table_dir, "paired_wilcoxon.tsv"))

cmp <- inner_join(paired, select(res_df, promoter, log2FoldChange), by = "promoter")
cat(sprintf("Spearman(median_delta, DESeq2 log2FC) = %.3f over %d promoters\n",
            cor(cmp$median_delta, cmp$log2FoldChange, method = "spearman",
                use = "pairwise"), nrow(cmp)))
```

# Session info

```{r session}
sessionInfo()
```
